The human visual system preserves the hierarchy of 2-dimensional pattern regularity

### 2021-01-21 Writeup/manuscript.tex

\begin{abstract}
A century of vision research has demonstrated that symmetry contributes to numerous domains of visual perception \cite{RN1311, RN1166,RN1682,RN1824}. In a 2-D image, the four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in 17 distinct ways. These 17 “wallpaper” groups \cite{RN1562,RN1563,RN1425} obey a hierarchy, determined by mathematical group theory, in which simpler groups are subgroups of more complex ones\cite{RN1711}. Here we probe representations of symmetries in wallpaper groups using two methods: (1) Steady-State Visual Evoked Potentials (SSVEPs) recorded using EEG and (2) symmetry detection thresholds measured psychophysically. We find that hierarchical relationships between the wallpaper groups are almost perfectly preserved in both behavior and response amplitudes in visual cortex. This remarkable consistency between the structure of symmetry representations and mathematical group theory, is likely generated over visual development, through implicit learning of regularities in the environment.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection \cite{RN1337}. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.

>>>>

\begin{abstract}Symmetries are present at many scales in images of natural scenes. A large body of literature has demonstrated contributions of symmetry to numerous domains of visual perception. The four fundamental symmetries, reflection, rotation, translation and glide reflection, can be combined in exactly 17 distinct ways. These \textit{wallpaper groups} represent the complete set of symmetries in 2D images. The goal of the current study is to provide a complete description of responses to symmetry in the human visual system, by collecting both brain and behavioral data using all 17 groups. This allows us to probe the hierarchy of complexity among wallpaper groups, in which simpler groups are subgroups of more complex ones. We find that this hierarchy is preserved almost perfectly in both behavior and brain activity: A multi-level Bayesian GLM indicates that for most of the 63 subgroup relationships, subgroups produce lower amplitude responses in visual cortex (posterior probability: >0.95 for 56 of 63) and require longer presentation durations to be reliably detected (posterior probability: >0.95 for 49 of 63). This systematic pattern is seen only in visual cortex and only in components of the brain response known to be symmetric-specific. Our results show that representations of symmetries in the human brain are precise and rich in detail, and that this precision is reflected in behavior. These findings expand our understanding of symmetry perception, and open up new avenues for research on how fine-grained representations of regular textures contribute to natural vision.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{S}ymmetries are present at many scales in images of natural scenes, due to a complex interplay of physical forces that govern pattern formation in nature. The importance of symmetry for visual perception has been known at least since the gestalt movement of the early 20th century. Since then, symmetry has been shown to contribute to the perception of shapes \cite{RN1311,RN1682}, scenes \cite{RN1824} and surface properties \cite{RN1166}, as well as the social process of mate selection. Most of this work has focused on mirror symmetry or \textit{reflection}, with much less attention being paid to the other fundamental symmetries: \textit{rotation}, \textit{translation} and \textit{glide reflection}. In the two spatial dimensions relevant for images, these four fundamental symmetries can be combined in 17 distinct ways, the “wallpaper” groups \cite{RN1562,RN1563,RN1425}. Previous work on a subset of four wallpaper groups used functional MRI to demonstrate that rotation symmetries in wallpapers elicit parametric responses in several areas in occipital cortex, beginning with visual area V3 \cite{RN1725}. This effect is also robust with electroencephalography (EEG), whether measured using Steady-State Visual Evoked Potentials (SSVEPs)\cite{RN1725} or event-related paradigms \cite{RN1959}. Here we extend this work by collecting SSVEPs and psychophysical data from human participants viewing the full set of wallpaper groups. We measure responses in visual cortex to 16 out of the 17 wallpaper groups, with the 17th serving as a control stimulus. Our goal is to provide a more complete picture of how wallpaper groups are represented in the human visual system.
